---
title: "EUTF data from AKVO"
output:
  md_document:
    variant: gfm
number_sections: yes
---
# EUTF for Africa: Data cleaning

**Script by:** [Kira Schacht](https://www.twitter.com/daten_drang)


# Files

| Name | Content |
|---|---|
| `cleaning/eutf_scrape.Rmd`	| The main R markdown script. Run in RStudio to reproduce this analysis. |
|`cleaning/data.RData` | The R Data file containing the finished datasets. |
|`data/...` | Data files |


# Data source

The data on EUTF projects comes from the AKVO monitoring platform. A full XML file of all projects linked to the EUTF is available [here](https://eutf.akvoapp.org/en/organisation/3394/).

The version used for this analysis was last updated on 17. December 2021.


# Create datasets

Here is a step-by-step-explanation of the code we used to create our analysis data sets. You can explore it yourself by opening `eutf_scrape.Rmd` in RStudio.

```{r setup, echo = FALSE}
#Setup

knitr::opts_chunk$set(echo = FALSE)
#load packages

knitr::opts_knit$set(root.dir = "/Users/schachtk/Documents/DW/xxx_EUTF/extern")

library(needs)
needs(tidyverse, XML)
```

## Read data

Read raw XML data from file `akvo-20211217-130832.xml`

```{r}

d = xmlParse("data/raw/akvo-20211217-130832.xml")

```


## Project info dataset

For every project (XML node `iati-activity`), create one row containing:

- IATI project ID
- project title
- project summary
- partner organizations
  - funding partners (in XML: `role=1`)
  - accountable partners (in XML: `role=2`)
  - extending partners (in XML: `role=3`)
  - implementing partners (in XML: `role=4`)
- project start and end dates (in XML: `activity-date type='2'` and `type='3'`)
- project budget
- recipient country or countries
- project location
- project topics according to the 4 thematic fields of the EUTF:
  1. Greater economic and employment opportunities
  2. Strengthening resilience
  3. Improved migration management
  4. Improved governance and conflict prevention
- OECD development aid CRS purpose codes
- URLS of linked documents
- related activities: parent projects, child  projects, sibling projects


```{r}
d_projects = xpathApply(d, "//iati-activity", function(node){
  data.frame(
    #IATI project ID
    iati_id = node %>% xpathSApply("./description[@avko:type='4']", xmlValue,
                                   namespaces = c("avko" = "http://akvo.org/iati-activities")) %>% ifelse(length(.)==0, NA, .),
    #Project title
    title = node[["title"]] %>% xmlValue %>% ifelse(length(.)==0, NA, .),
    #summary
    summary = node %>% xpathSApply("./description[@avko:type='5']", xmlValue,
                                   namespaces = c("avko" = "http://akvo.org/iati-activities")) %>% ifelse(length(.)==0, NA, .),
    #partner organizations
    partner_funding = node %>% xpathSApply("./participating-org[@role='1']", xmlValue) %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, NA, .),
    partner_accountable = node %>% xpathSApply("./participating-org[@role='2']", xmlValue) %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, NA, .),
    partner_extending = node %>% xpathSApply("./participating-org[@role='3']", xmlValue) %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, NA, .),
    partner_implementing = node %>% xpathSApply("./participating-org[@role='4']", xmlValue) %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, NA, .),
    #project start and end dates
    date_start = node %>% xpathSApply("./activity-date[@type='2']", xmlGetAttr, "iso-date") %>% ifelse(length(.)==0, NA, .) %>% as.Date,
    date_end = node %>% xpathSApply("./activity-date[@type='3']", xmlGetAttr, "iso-date") %>% ifelse(length(.)==0, NA, .) %>% as.Date,
    #project budget
    budget = node %>% xpathSApply("./budget", xmlValue) %>% as.numeric %>% sum %>% ifelse(length(.)==0, NA, .),
    #recipient countries
    recipient_country = node %>% xpathSApply("./recipient-country", xmlGetAttr, "code") %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, "", .),
    #project location
    location = node %>% xpathSApply("./location/name/narrative", xmlValue) %>% paste(., collapse = ", ") %>% ifelse(length(.)==0, "", .),
    location_code = node %>% xpathSApply("./location/administrative", xmlGetAttr, "code") %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, "", .),
    #project topics
    topics = node %>% xpathSApply("./sector", xmlGetAttr, "code") %>% `[`(grepl("TFP",.)) %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, "", .),
    #OECD development aid CRS purpose codes
    oecd_purpose = node %>% xpathSApply("./sector[vocabulary='1']", xmlGetAttr, "code") %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, "", .),
    #linked documents
    links = node %>% xpathSApply("./document-link", xmlGetAttr, "url") %>% `[`(grepl("pdf|doc",.)) %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, "", .),
    #related activities:
    project_parent = node %>% xpathSApply("./related-activity[@type='1']", xmlGetAttr, "ref") %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, NA, .),
    project_child = node %>% xpathSApply("./related-activity[@type='2']", xmlGetAttr, "ref") %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, NA, .),
    project_sibling = node %>% xpathSApply("./related-activity[@type='3']", xmlGetAttr, "ref") %>%
      paste(., collapse = ", ") %>% ifelse(length(.)==0, NA, .),
    stringsAsFactors = FALSE
  )
}) %>%
  #combine into one data frame
  bind_rows() %>%
  #clean IATI IDs
  mutate(across(c(17:19), gsub, pattern = "XI[-_]IAT[IA][-_]EC[-_]DEVCO[-_]|XI-IATI-EC_NEAR_", replacement = "", ignore.case = TRUE)) %>%
  #remove NULL values
  mutate(across(c(recipient_country, location_code), gsub, pattern = "NULL(, )?", replacement = "")) %>% 
  #transform empty values to NA
  mutate(across(everything(), na_if, "")) %>% 
  #separate project titles into prefix (DEC or CRT) and descriptive title
  separate(title, into = c("prefix","title"), sep = "(?<=^[CTRDEC]{3}) ?-? ?", fill = "left", extra = "warn")

#manually fix missing IATI ID
d_projects$iati_id = gsub("CTR - Mission d'appui aux acteurs du programme de renforcement de la résilience à l'insécurité alimentaire et nutritionnelle au Nord mali (Key)","T05-EUTF-SAH-ML-01-04", d_projects$iati_id, fixed = T)

#write to file
write.csv(d_projects, "data/processed/EUTF_projects_akvo.csv", row.names = F, na = "")

d_projects[c(1:3,8:12)] %>% arrange(-budget) %>% head(10) %>% knitr::kable()
```


## Results list

For every project, create one row for every result indicator containing:

- IATI project ID
- result type: Reporting category of results, e.g. `EUTF Common Output Indicators`
- title of indicator
- reporting period start and end dates
- target value
- actual value

```{r}

d_results = xpathApply(d, "//iati-activity", function(node){
  # note ID for every project
  iati_id = node %>% xpathSApply("./description[@avko:type='4']", xmlValue,
                                 namespaces = c("avko" = "http://akvo.org/iati-activities")) %>% ifelse(length(.)==0, NA, .)
  #loop through all "result" nodes
  node %>% xpathApply("./result", function(result){
    
    # note result type for each
    result_type = result %>% xpathSApply("./title", xmlValue) %>% ifelse(length(.)==0, NA, .) %>% as.character()
    
    #loop through all "indicators" within result nodes
    result %>% xpathApply("./indicator", function(indicator){
      
      data.frame(
        #title of indicator
        title = indicator %>% xpathSApply("./title", xmlValue) %>% paste(., collapse = ";;") %>% ifelse(length(.)==0, "", .),
        # reporting period start and end
        period_start = indicator %>% xpathSApply("./period/period-start", xmlGetAttr, "iso-date") %>%
          ifelse(length(.)==0, "", .) %>% as.Date,
        period_end = indicator %>% xpathSApply("./period/period-end", xmlGetAttr, "iso-date") %>%
          ifelse(length(.)==0, "", .) %>% as.Date,
        # target value
        target = indicator %>% xpathSApply("./period/target", xmlGetAttr, "value") %>%
          paste(., collapse = ";;") %>% ifelse(length(.)==0, "", .),
        # actual value
        actual = indicator %>% xpathSApply("./period/actual[not(./dimension)]", xmlGetAttr, "value") %>%
          paste(., collapse = ";;") %>% ifelse(length(.)==0, "", .),
        stringsAsFactors = FALSE
      )
    }) %>%
      # combine into data frame, add result type at the front
      bind_rows() %>% mutate(result_type = result_type, .before = everything())
      
  }) %>%
    # combine into data frame, add project ID at the front
    bind_rows() %>% mutate(iati_id = iati_id %>% as.character, .before = everything())
}) %>%
  bind_rows() # combine into data frame

write.csv(d_results, "data/processed/EUTF_akvo_results.csv", na = "") #write to file


d_results %>% head(10) %>% knitr::kable()
```
## Partner organizations list

Create a list of every partner organization witht he following information:

- name
- AKVO type: a numerical value describing the type of organization
- the number of occurrences in each role:
  - funding partner (in XML: `role=1`)
  - accountable partner (in XML: `role=2`)
  - extending partner (in XML: `role=3`)
  - implementing partner (in XML: `role=4`)

```{r}
#partners list
d_orgs = data.frame(
    name = xpathSApply(d, "//iati-activity/participating-org", xmlValue),
    role = xpathSApply(d, "//iati-activity/participating-org", xmlGetAttr, "role"),
    type = xpathSApply(d, "//iati-activity/participating-org", xmlGetAttr, "type"),
    stringsAsFactors = FALSE
  ) %>% 
  # count occurrences by role
  group_by(name, type, role) %>% summarise(n = n()) %>% 
  # create one column for every role
  pivot_wider(c(name, type), names_from = role, names_sort = TRUE, values_from = n, values_fill = 0) %>% 
  rename(funding = `1`, accountable = `2`, extending = `3`, implementing = `4`)

#write to file
write_csv(d_orgs, "data/processed/EUTF_akvo_orgs.csv", na = "")

d_orgs %>% arrange(-implementing) %>% head(10) %>% knitr::kable()
```
## Recipient country list

For every project, list recipient countries and, if available, the `percentage` attributed to them, as follows:

- IATI project ID
- project budget
- ISO alpha-2 country code of recipient country
- percentage attributed to recipient country

This can be used to estimate the budget share of regional projects attributable to each recipient country.

```{r}
d_recipients = xpathApply(d, "//iati-activity", function(node){
  
  iati_id = node %>% xpathSApply("./description[@avko:type='4']", xmlValue,
                                 namespaces = c("avko" = "http://akvo.org/iati-activities")) %>% ifelse(length(.)==0, NA, .)
  budget = node %>% xpathSApply("./budget", xmlValue) %>% as.numeric %>% sum %>% ifelse(length(.)==0, NA, .)
  
  node %>% xpathApply("./recipient-country", function(country){
    
    data.frame(
      recipient_country = country %>% xpathSApply(".", xmlGetAttr, "code") %>% paste(., collapse = ";;") %>% ifelse(length(.)==0, NA, .),
      recipient_percentage = country %>% xpathSApply(".", xmlGetAttr, "percentage") %>% paste(., collapse = ";;") %>% ifelse(length(.)==0, NA, .),
      stringsAsFactors = FALSE
      )
    }) %>% bind_rows() %>%
    mutate(across(everything(), gsub, pattern = "NULL(, )?", replacement = NA)) %>% 
    mutate(iati_id = as.character(iati_id), budget = budget, .before = everything())
}) %>% bind_rows()

write.csv(d_recipients, "data/processed/EUTF_akvo_recipients.csv", na = "")

d_recipients %>% head(10) %>% knitr::kable()
```




## List of discontinued projects

To find projects that might have been discontinued, we search for projects referred to as "related activities" in the database that don't have their own separate entry.

1. list all IATI IDs in separate `iati-activity`entries
2. list all IATI IDs referred to in `related-activity`tags
3. list which entries of 2. don't appear in 1.

```{r}
iati_id = d %>% xpathSApply("//iati-activity/description[@avko:type='4']", xmlValue,
                            namespaces = c("avko" = "http://akvo.org/iati-activities"))

all_iati = d %>% xpathSApply("//iati-activity/related-activity", xmlGetAttr, "ref") %>% unique %>%
  gsub(pattern = "XI[-_]IAT[IA][-_]EC[-_]DEVCO[-_]|XI-IATI-EC_NEAR_", replacement = "", ignore.case = TRUE)

all_iati[!(all_iati %in% iati_id)] %>% sort %>% paste(collapse = ", ")
```

```{r}
save.image("cleaning/data.RData")
```

